---
title: "Anti Xa monitoring in ECMO "
subtitle: "⚔<br/>Journal Club"
author: "Teddy Tun Win HLA <br/> Senior Clinical Fellow in Intensive Care Medicine"
institute: "University College London Hospital "
date: "2022/02/01 (updated: `r Sys.Date()`)"
output: 
  xaringan::moon_reader:
    self_contained : true
    css: [default, metropolis, metropolis-fonts]
    lib_dir : libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# PLAN

<br>
<br>
- Brief background
- Current Findings
- Further Steps
- Discussion
<br>

```{r}
library(knitr)
library(ggplot2)
source("datatidy.R")
```

--


<br>

- Estimated time : **30 mins.**
---
# Anticoagulation in ECMO 

- Extra corporeal therapies - > anticoagulation. 
- GSTT - a large volume severe respiratory failure centre providing predominant VV-ECMO service. 
- Bleeding and thrombotic complications during ECMO are NOT uncommon
- __Multiple options__ : Activated Clotting Time (ACT), Activated Partial Thromboplastic time (aPTT),anti-factor Xa (Anti-Xa),prothrombin time and internationalised normal ratio (PT/INR)
--
<br>
<br>
<br>
- __apTT vs aXa__ ? 
- factor VIII, fibrinogen, lupus anticoagulant, acute phase reactants affect apTT
- Research challenging : due to low volume.
- small signal aXa may be a superior correlate - paediatric ecmo<sup>1</sup> 

.footnote[
[1] Paeds ecmo aXa
[link](https://pubmed.ncbi.nlm.nih.gov/24335992/)
]
--

---
# Is anti-Xa monitoring better?

### Methodology
- single centre retrospective cohort review with electronic health records interrogation
- ?Ethics
### Opportunities
- Large volume centre
- COVID viral pneumonia and previous flu pandemic : viral aetiology
### Clinical Outcomes
- Time to first therapeutic range, Time in therapeutic range
- Bleeding and Thrombotic Complications (e.g., Time to 1st complication)
### Non Clinical Outcomes
- Frequency of prescription change (reduced workload on clinical team)
---
# Early Findings
n = 201 (Flu : 80, COVID
Wave 1 : 59, COVID Wave 2 : 62)
```{r plot1, echo=FALSE,message = FALSE,warning = FALSE, out.width = "500px",fig.retina=2}

ggplot(pt, aes(x=admissiondate,color = cohort))+
               geom_bar(binwidth = 86400)

```

---
# Missing Data
---
# Next Steps
---
# Discussion
---
# Feedback
---