---
title: "Anti Xa monitoring in ECMO: a preliminary analysis "
subtitle: "⚔<br/>GSTT Journal Club"
author: "Teddy Tun Win HLA <br/> Senior Clinical Fellow in Intensive Care Medicine"
institute: "University College London Hospital "
date: "2022/02/09 (updated: `r Sys.Date()`)"
output: 
  xaringan::moon_reader:
    self_contained : true
    css: [default, metropolis, metropolis-fonts]
    lib_dir : libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# PLAN

<br>
<br>
- Brief background
- Current Findings
- Further Steps
- Discussion
<br>
<br>
- **DISCLAIMER** - My work is in progress and is incomplete. 
--

```{r initialising, include=FALSE}
library(knitr)
library(ggplot2)
library(plotly)
library(tidyverse)
library(survminer)
library(survival)
library(kableExtra)
library(emo)
source("datatidy.R")
```
--

<br>

- Estimated time : **30 mins.**
---
# Anticoagulation in ECMO 

- Extra corporeal therapies - > anticoagulation. 
- GSTT - a large volume severe respiratory failure centre providing predominant VV-ECMO service. 
- Bleeding and thrombotic complications during ECMO are NOT uncommon. 
- __Multiple options__ : Activated Clotting Time (ACT), Activated Partial Thromboplastic time (aPTT),anti-factor Xa (Anti-Xa),prothrombin time and internationalised normal ratio (PT/INR)
--
<br>
<br>
<br>
- __apTT vs aXa__ ? 
- factor VIII, fibrinogen, lupus anticoagulant, acute phase reactants affect apTT
- Research challenging : due to low volume.
- small signal aXa may be a superior correlate - paediatric ecmo<sup>1</sup> 
- theoretical : intra-test variability reduced. 

.footnote[
[1] Paeds ecmo aXa
[link](https://pubmed.ncbi.nlm.nih.gov/24335992/)
]


---
# Is anti-Xa monitoring better?

- Definition of 'better' 

> Better **efficacy** -- results in improved time in therapeutic range?

--

> Better **patient** outcomes -- safer/less complications ? 

--


> Better **provider** outcomes -- less monitoring/circuit change/dose change?

---
# Evaluation of anti-Xa monitoring: a pilot study
### Methodology
- single centre retrospective cohort review with electronic health records interrogation
- ?Ethics (service evaluation +/-)

--

### Opportunities
- Large volume centre
- COVID viral pneumonia(wave1&2) and previous flu pandemic : viral aetiology
- Timeframe : 01-01-2015 to 01-03-2021

--

### Clinical and Non-Clinical Outcomes
- Time to first therapeutic range, Time in therapeutic range
- Bleeding and Thrombotic Complications (e.g., Time to 1st complication)
- Frequency of prescription change (reduced workload on clinical team)

---
# Note on Data

Q1. Are there other anticoag used?

> A1. No (only heparin)


Q2. Ecmo run time used as gold standard?

> A2. Yes


Q3. No repeat ecmo runs?

> A3. Yes

Q4. Peak ecmo cannulations - 3 in 24 hours?

> A4. Yes covid. 


Q5. Data source?

> A5. EHR heparin prescription, anti-Xa and APTTR results, complications.

---
# Heparin Prescription and Protocol

Heparin administered in 4 different formats 
- Heparin units/hour
- Heparin ECMO Xa infusion units/kg/hour
- Heparin IV Injection units
- Heparin RRT Inf 2000 iu units
- Heparin systemic inf 2000 units units/hour

--

-Thrombotic complications NOT available / included in this analysis.

-Bleeding complications extracted from EHR. 

---
# Description of Patient Population 

- Sensecheck 

```{r demographics table, echo=FALSE,message = FALSE}
pt %>% 
        select(cohort,survivedecmo,ecmorunt) %>%
        group_by(cohort) %>% 
        summarise(
          no_pts = n(),
          deaths_ecmo = sum(survivedecmo == "no"),
          crude_mortality_rate = round(deaths_ecmo/no_pts,digits = 2),
          mean_ecmot = round(mean(ecmorunt),digits = 2),
          median_ecmot = median(ecmorunt),
          min_ecmot = min(ecmorunt)
                  ) %>%
        kbl(format= "html",full_width = F) %>% 
          kable_styling()
        
```

---
# Admission Frequency of ECMO patients

```{r admission_f_plot, echo = FALSE, dpi = 300, fig.width = 6, fig.height = 4}

p2 <- ggplot(data = pt, aes(x = as.Date(admissiondate),colour = cohort)) +
      geom_bar()+
      scale_x_date(date_breaks = "6 months",date_minor_breaks = "3 months",date_labels = "%b %Y")+
      theme_minimal()+
      theme(axis.text.x = element_text(angle = 45,hjust = 1 ))+
      labs (x = "Time", title = "ECMO admissions across study period", y = "Admissions per day")
     

p2
```

---
# Age Distribution and Cohort Outcomes

.pull-left[

```{r age,echo=F,dpi=300,fig.width=3,fig.height=3}
p4 <- ggplot(pt, aes(x = cohort, y= age, fill = cohort)) + 
        geom_boxplot() +
        geom_jitter(alpha = 0.4,width = 0.2, aes(colour = cohort))+
        geom_hline(linetype = 2 , yintercept = 46) +
        labs(title = "Age distribution across cohorts",subtitle = "Median age 46, Kruskal-Wallis p 0.2",y= "Age / years")+
        theme_minimal()

p4

```

]

--

.pull-right[

```{r cohort,echo=F,dpi=300,fig.width=3,fig.height=3}
p5 <- ggplot(data = pt, aes(x = cohort, fill = survivedecmo))+
    geom_bar(position = "stack") + 
    theme_minimal()+
    labs(x = "cohort", y = "Number of Patients",title = "Patients who survived ECMO per cohort", subtitle = "No statistical difference")

p5
```


]

Age : median age is very similar across cohorts.

Mortality : wave 2 is higher but not statistically sig.

---
# Frequency of ECMO Run Times per cohorts

```{r echo = FALSE, dpi = 300,fig.width = 6, fig.height = 3.5 }
p1 <- ggplot(data = pt, aes(x=as.numeric(ecmorunt), colour = cohort)) +
   geom_freqpoly(binwidth = 5 ) +
   scale_x_continuous(breaks = seq(0,90,by=10))+
   theme_minimal()+
   labs (x = "ECMO run time / days", y = "Patients", title = "Frequency of ECMO run times per cohorts")

p1
```


---
# Time on ECMO to outcome

```{r echo = FALSE, dpi = 300, fig.width = 6, fig.height = 4}
ps <- survfit (Surv(ecmorunt,survivedecmo == "no")~cohort, data = pt)

p6 <- ggsurvplot(
        ps,
        pval = TRUE,
        risk.table = F, 
        risk.table.col = "cohort",
        risk.table.height = 0.3,
        surv.median.line = "hv") + labs (title = "Time on ECMO to outcome",x = "ECMO run time / days",y = "% on ECMO")

p6 

```

---
# Ecmo run time risk table

```{r echo= F, dpi = 300, fig.width= 6, fig.height=4}

t1 <- ggsurvtable(ps,color = "cohort", data = pt,title = "% on ECMO over time in days")
t1$risk.table

```

---
# Anticoagulation

- **NO GOLD** standard monitoring profile.


- Several solutions


> Time to first therapeutic range

> Time in therapeutic range (TTR)


- Most of the information from warfarin / dOAC monitoring.
---

# Time in therapeutic range defintions


| Method       | *Principle*     | Rosendaal            | Traditional            | Cross-sectional                       |
|--------------|---------------|----------------------|------------------------|---------------------------------------|
| Numerator    | *Time in Range* | No of days in range  | No of in-range visits  | No of patients in range on last visit |
| Denominator  | *Total Time*    | Total Monitored Days | Total no of **visits** | Total no of **patients**              |
 

- Cross sectional method is not applicable to our population. 

- Traditional method assumes INR is static and binary (in or out of range).


- Rosendaal method(1993) assumes INR as dynamic but a **linear** progression between visits.
- Most studies do not mention method for calculating TTR.

---
# Raw data visualisation
<img src = "p9.png" align = "middle" size="280px" />
---
# Visualisation per cohort
<img src = "p12.png" align = "middle" size = "400px" />
---
# Time to first therapeutic range

- group assignment problem?

- 1st of November 2019 : **before** apttr monitoring and **after** axa monitoring.

- However, most patients have both apttr and axa sent.

> Likely early trial of anti Xa pre 1/Nov/2019.

- Even after 1st Nov 2019, axa monitoring may not be set (?)

- Some clinical team prefer apttr (?)

> large value of missing AXA value after 1/Nov/2019


---
# Missingness 1/2

- ** B I G ** topic in EHR analysis.

- Implicit vs. Explicit missing data.

- Explicit : declared as 'NONE' or 'NA' in the database.
- Implicit : simply not present. 

<img src= "img/ex.png" width = "25%" align = "right"/>

<img src = "img/im.png" width = "25%" align = "left" />


 Water?

<br>
<br>

|   | No of patients | Blood results | Heparin Prescription | Complications |
|---|----------------|---------------|----------------------|---------------|
| n | 201            | 201           | 185                  | 63            |

---
# Missingness 2/2 

Truth  1: Patient X, 3 days on VV-ECMO, 3 anti-Xa checks, 6 APTTR checks 

Dataframe

| Patient | Timestamp    | anti Xa value | apttr value |
|---------|--------------|---------------|-------------|
| X       | day 1 AM     | 0.5           | 1.5         |
| X       | day 1 PM     | NA            | 1.7         |
| X       | day 2 AM     | 0.6           | 1.4         |
| X       | ... day 3 PM | NA            | 1.5         |

vs. 

Truth : Patient X predominantly monitored on apttr but have occassional AXA checks

---
# Next Steps WITH current data- 1/2 

- Complete my analysis.`r emo::ji("smile")`

- Missing data needs cross-checking and perhaps re-extracting. e.g., 2 blood values with same time stamp.

- Need input from clinician (i.e., you) on what is appropriate. 

- Need further information on an existing patient popn 

> e.g., thrombotic complications, complete haemorrhagic complications, 

- Other approaches (e.g., Bland-Altman) to interrogate bias in assay measurements.

---

# Next Steps WITH MORE data - 2/2 

- Further variables on patient information : 
    - ethnicity / gender / renal function / other Rx e.g., aspirin / Past medical history
    - SOFA score at admission / Hb/ plts
    - no of red cells / platelets transfused 

- Circuit information?
    - Historical circuit stock e.g., MAHAM/NO doped circuits?
    - circuit change frequencies

- Proposal to expand patient popn to e.g., 31st Dec 2021

---
# Ambition and Discussion


- Expand to ecmo popn OR general ICU popn
> Risk : minimal as easily scaleable, potentially messier data with more heterogeneous popn.
> Benefits : larger volume. 

- Inference of variables associated with outcomes
> Risks vs. benefits as per above. 

- Prediction of 'optimal' dosing (e.g., via app )
> Risks : prediction / validation / app development pipeline

- Validation with other centre data - e.g., Brompton ? 
> Risks : access to data
> Benefits : potentially multi-centre data.

---
# Acknowledgements and Feedback

- Many thanks to Dr Andy Retter and Dr Barney Sanderson for data extraction and guidance.

- I am most grateful for your time and feedback.

- THANK YOU
