---
title: "Anti Xa Monitoring : A Preliminary Analysis"
author: "teddyhla"
date: "25/01/2022"
output: 
  html_document:
    number_sections : TRUE
    toc : true
    toc_float : true
    theme : flatly

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warnings = F)
#this is a global chunk options.
#for release - echo should be FALSE which means results will only appear. messages and warnings should also be false. 
```

<style>
@import url('https://fonts.googleapis.com/css2?family=Assistants&display=swap" rel="stylesheet');
body {
font-family: 'Assistant', sans-serif;
}
</style>

# TO-DO
- explore missingess and label  'missingess at random' vs'missingess not at random' vs. missingess at completely random'.
- explore no of lab draws 
- explore those who dont reach targets and mortality
- explore icu survivability vs. ecmo survivability i.e., are patients dying on ecmo contributing to icu mortality or is it in post ecmo phase
- explore time to in range vs. time to mortality
- explore rosenfeld method and need definition for method.
- are the patients who never got into range those who are never anticoagulated? 

# Introduction

- Any extra-corporeal therapy requires some form of anticoagulation to reduce thrombotic complications. 
- Anticoagulation in critically unwell patients is challenging with rapid changes in physiology, biochemistry and disease progress.
- Several anticoagulants are available but intravenous unfractionated heparin is most commonly used.
- Modalities to monitor heparin exist from long-term monitoring (e.g., INR for warfarin monitoring) or from static coagulation tests.[cite]
- APTT test - activated partial thrombplastin time (aPTT) assay is a standard modality, available since the 1950's.
- APTT test is susceptible to interference from factor VIII or fibrinogen concentrations.
- Anti Xa assay - is thus a more attractive option as is not susceptible to factor deficiencies except anti-thrombin deficiency.
- Whilst anti Xa assay is recommended for monitoring by American College of Chest Physicians, College of American Pathologists, there is no definitive study showing the superiority of anti-Xa. 


# Objectives

- Does the use of anti-Xa monitoring results in 'better' outcomes?
- Definition of 'better' :

> Better efficacy - results in improved time in therapeutic range?

> Better patient outcomes - results in safer (i.e., less complications) in patients?

> Better provider outcomes - less monitoring frequency and drug dosage changes, potential for time-saving, cost(circuit), safety?


# Ambition

- monitor all critically unwell patients with Anti-Xa ?

- development of an applet to guide dosage and monitoring?

- identification of variables associated with complications / worse outcomes? 


# Pilot: Dose the use of Anti-Xa monitoring in VV-ECMO patients improve outcomes?

- Explicit Inclusion criteria : all patients on VV-ECMO for viral respiratory failure from 01-01-2015 to 01-03-2021.

> Implicit Assumptions : all adults, predominant single organ failure of similar viral aetiology. 

> Traditional Exclusion critiera : severe liver diseases, haematological malignancy implicitly excluded. 

> Note : Pregnant patients in our cohort.

- Timeframe : 1st November 2019 GSTT ECMO service switches to anti-Xa monitoring. 

- Methodology : Retrospective cohort study of all eligible patients, electronic health records interrogation to yield

1. patient's on ECMO run time 
2. Measured aPTTR and AXA levels whilst patients on ECMO
3. Anticoagulant prescriptions whilst on ECMO
4. Complications whilst on ECMO 


# Source of Data

## Cohorts 

Three cohorts of severe respiratory failure patients are identified :- 

- wave 1 of covid - 19, defined as between 01-03-2020 to 31-08-2020 inclusive.

- wave 2 of covid - 19, defined as between 01-09-2020 to 01-03-2021 inclusive.

- historical flu, defined as 01-01-2015 and 31-12-2019 inclusive. 

## Questions regarding data.

- What about non covid and non flu patients in this period? isnt it better to capture 01-01-2015 to 01-01-2022 ? 
- Ans : better to focus on viral aetiology. 

- What about large gaps - is it true?
- Ans : as per above 

- What about lmwh / flolan / argotraban ? 
- Ans : only heparin was used 

- Use ecmo run time as gold standard? 
- Ans : agreed 

## Sensecheck 

- No repeat admissions?
- Ans : true

- Peak 3 ecmo admission in 24 hour ?
- Ans : true

## Patient identification
- ecmo blood pump speed was calculated from min to max dates and subtracted to get ecmo run times and patients

## Heparin administration

Heparin was administered in 4 different formats 

- heparin units/hr
- heparin ecmo xa iv infusion units/kg/hr
- heparin iv injection units
- heprin rrt inf 2000 iu units
- heparin systemic inf 2000 units units/hr

- Ans : agree this is the complexity of heparin prescription in ecmo.

### Heparin Protocol

Intravenous unfractionated heparin infusion (UFH) -  50 IU/kg as a bolus at the
time of cannulation with 2,500 units in the priming fluid for the circuit unless contraindicated
due to bleeding.


Prior to 1st November 2019, UFH infusions were monitored by activated partial thromboplastin
time ratio (APTTr) with a target level of 1.5-2.0 if evidence of no thrombosis and haemorrhage
and 2.0-2.5 if a thrombotic event.


From 1st November 2019 to present, anti-Xa activity was used for monitoring targeting levels
of 0.3 - 0.7 unit/mL if no evidence of thrombosis or haemorrhage were present and 0.6 - 1.0
unit/mL if there was a thrombotic ev

Weight is also extracted to calculate units/hr/hr 

## Coagulation lab results
- anti Xa and APTTR results were extracted

## Complications 

- Bleeding scores are extracted from EHR. ?Is this **ELSO** definitions?
- Thrombotic complications **NOT** available.


# Data Analysis

## Initialising 
### Load dependencies

```{r loading_libs}
library(ggplot2)
library(cowplot)
library(survminer)
library(survival)
library(plotly)
library(kableExtra)
library(tidylog)
```

### Source the cleaned files
```{r loading_source, message=TRUE,warning=FALSE}
source("scripts/0-importclean.R")
```


### Methods of Statistical Analysis

### Type of Reagent and Assay used

## Description of Patient Population 

### Gross Statistics 
```{r baseline_table_01}
pt %>% 
        select(cohort,survivedecmo,ecmorunt) %>%
        group_by(cohort) %>% 
        summarise(
          no_pts = n(),
          no_deaths_ecmo = sum(survivedecmo == "no"),
          crude_mortality_rate = round(no_deaths_ecmo/no_pts,digits = 2),
          mean_ecmorunt = round(mean(ecmorunt),digits = 2),
          median_ecmorunt = median(ecmorunt),
          min_ecmorunt = min(ecmorunt)
                  ) %>%
        kbl() %>% 
        kable_material_dark()
        
```


### Admission Frequency of ECMO patients

```{r admission_frequency_cohort}
p2 <- ggplot(data = pt, aes(x = as.Date(admissiondate),colour = cohort)) +
      geom_bar()+
      scale_x_date(date_breaks = "6 months",date_minor_breaks = "3 months",date_labels = "%b %Y")+
      theme_minimal()+
      theme(axis.text.x = element_text(angle = 45,hjust = 1 ))+
      labs (x = "Time", title = "ECMO admissions across study period", y = "Admissions per day")
     

p2
```


### Distribution of ECMO run times.

```{r hist_ecmo_runtimes}
p1 <- ggplot(data = pt, aes(x=as.numeric(ecmorunt), colour = cohort)) +
   geom_freqpoly(binwidth = 5 ) +
   scale_x_continuous(breaks = seq(0,90,by=10))+
   theme_minimal()+
   labs (x = "ECMO run time / days", y = "Patients", title = "Frequency of ECMO run times per cohorts")

p1
```

### Dose age and cohort effect ECMO survival? 
```{r Age distribution across cohorts}

p4 <- ggplot(pt, aes(x = cohort, y= age, fill = cohort)) + 
        geom_boxplot() +
        geom_jitter(alpha = 0.4,width = 0.2, aes(colour = cohort))+
        geom_hline(linetype = 2 , yintercept = 46) +
        labs(title = "Age distribution across cohorts",subtitle = "Median age 46, Kruskal-Wallis p 0.2",y= "Age / years")+
        theme_minimal()

p4
```

```{r cohort_outcomes}
p5 <- ggplot(data = pt, aes(x = cohort, fill = survivedecmo))+
    geom_bar(position = "stack") + 
    theme_minimal()+
    labs(x = "Cohorts", y = "Number of Patients",title = "Patients who survived ECMO per cohort", subtitle = "No statistical difference")

p5 
```


### Time on ECMO to outcome 

```{r survival data}
ps <- survfit (Surv(ecmorunt,survivedecmo == "no")~cohort, data = pt)

p6 <- ggsurvplot(
        ps,
        pval = TRUE,
        risk.table = F, 
        risk.table.col = "cohort",
        risk.table.height = 0.3,
        surv.median.line = "hv") + labs (title = "Time on ECMO to outcome",x = "ECMO run time / days",y = "% on ECMO")

p6 
```

### Risk Table for ECMO Run Times

```{r}
t1 <- ggsurvtable(ps,color = "cohort", data = pt,title = "% on ECMO over time in days")
t1$risk.table

```

## Anticoagulation Profiles

- **No** gold standard monitoring profiles.

> PROPOSED SOLUTIONS 

> - Time to first therapeutic range
> - Time in therapeutic range (TTR)

### Note on TTR 

- No gold standard measure.
- Studied extensively on warfarin/ DOAC in AF 
- Several methods

| Method       | *Principle*     | Rosendaal            | Traditional            | Cross-sectional                       |
|--------------|---------------|----------------------|------------------------|---------------------------------------|
| Numerator    | *Time in Range* | No of days in range  | No of in-range visits  | No of patients in range on last visit |
| Denominator  | *Total Time*    | Total Monitored Days | Total no of **visits** | Total no of **patients**              |
 

- Cross sectional method is not applicable to our population. 
- Traditional method assumes INR is static and binary (in or out of range).
- Rosendaal method(1993) assumes INR as dynamic but a **linear** progression between visits.
- Most studies do not mention method for calculating TTR.
- would trial a rolling average model.

### Exploration 

- NOTE :Calculations adjusted for patients with PE (n = 2 , who have different threshold.)
```{r exploration_anticoagulation}
clab <- clab %>% 
  group_by(mrn) %>% 
  mutate(ecmo_day = round(difftime(dtm,datecannulated,units = "days")),ecmo_min = dtm - datecannulated ) %>%
  ungroup()

clab$ecmo_day <- as.factor(clab$ecmo_day)

```

```{r range_axa_apttr}
ylim1 <- boxplot.stats(clab$axa)$stats[c(1,5)]
ylim2 <- boxplot.stats(clab$apttr)$stats[c(1,5)]

p7<-ggplot(data = clab, aes(x = ecmo_day, y = axa)) +
    geom_boxplot(outlier.shape = NA)+
    geom_hline(yintercept = c(0.3,0.7),colour = "blue")+
    coord_cartesian(ylim = ylim1 *1.05)+
    theme(axis.text.x = element_text(angle = 45))+
    theme_minimal()+
    labs(title= "Anti Xa ranges for each ECMO run day",
         x = "Day on Ecmo",y ="anti-Xa levels",
         subtitle = "Normal Range 0.3 to 0.7, VTE Range 0.6 to 1.0")

p8<- ggplot(data = clab, aes(x = ecmo_day, y = apttr)) +
        geom_boxplot(outlier.shape = NA)+
        geom_hline(yintercept = c(1.5,2.0),colour = "blue")+
        coord_cartesian(ylim = ylim2 *1.05)+
        theme(axis.text.x = element_text(angle = 45))+
        theme_minimal()+
        labs(title= "APTTR ranges for each ECMO run day",
         x = "Day on Ecmo",y ="APTTR valuess",
          subtitle = "Normal Range 1.5 to 2.0, VTE Range 2.0 to 2.5")

p9 <- plot_grid(p7,p8)
p9
```


```{r exploring_by_cohort}

p10 <- p7 +facet_wrap(~cohort)
p11 <- p8 + facet_wrap(~cohort)
p12 <- plot_grid(p10,p11)
p12
```


```{r duplicated_dtm_within_each_mrn}

#for each patient, we explore if there are duplicated time stamps.

#clab %>% group_by(mrn,dtm) %>% filter(duplicated(dtm))

```

6584394E

2019-01-21 03:30:00
has 2 lab values on same time stamps.

5743542Q
2015-01-02 11:53:00 has 4 different values of apttr.
There is 43 patients.

### Group assignment

- First problem is that group assignment 
- anti Xa monitoring is officially commenced on 1st November 2019.
- So assigning patient's 

### Let's try and find first time axa is in range.

# histogram of ecmoruntimes showed a median run time of 13 days.
#histogram of number of lab results showed a median of 34 days.
# thus dividing median by median results = 34/13 = 2.6 blood draws per day.


```{r exploring_axa_never_in_range}
#note AXA is started to use in 1st Nov 2019.
cin <- clab %>% filter(axarange == "in" & cohort != "Flu") %>% group_by(mrn)%>%ungroup()

cin2 <- clab %>% filter(mrn %in% setdiff(pt$mrn,unique(cin$mrn)))
cin2 <- cin2[cin2$cohort != "Flu",]
#this shows that 14 patients in wave 1 and wave 2 COVID cohorts never were used to monitor AXA. OR IF they did, they never reach the range. cin2 is a df showing this

#may be we need to consider tolerances.
```

```{r create_timettr}

dfaxa <- clab %>% filter(axarange == "in" & datecannulated > as.Date("2019-12-01")) %>% group_by(mrn) %>% slice_min(dtm) %>% ungroup()
dfapt <-clab %>% filter(apttrrange == "in" & datecannulated < as.Date("2019-12-01")) %>% group_by(mrn) %>% slice_min(dtm) %>% ungroup()
```


```{r}
#lets' check the code works
#j<- c("1103375H","4946122T","4946122T")
#craxa %>% filter (mrn %in% j) %>% View()

dfaxa$faxa <- dfaxa$dtm -dfaxa$datecannulated
dfapt$fapt <- dfapt$dtm - dfapt$datecannulated

dfaxa$faxa <- as.numeric(dfaxa$faxa)
dfapt$fapt <- as.numeric(dfapt$fapt)

wilcox.test(dfaxa$faxa,dfapt$fapt)

#but note that patient's 6 patients are not really have axa monitoring. 

```

```{r now}
cplab <-clab
cplab$group <- ifelse(cplab$datecannulated < as.Date("2019-12-01"),"apt","axa")
cplab$group <- as.factor(cplab$group)
#this is becuase people dont actually start using it until 2019-12-01
#however there are other cases also in wave 1 and 2 who did not receive axa monitoring.
#so any patients who have 70% or more missing from axa assigned as apttr
reassign_group_70p <- c("6950251Z","6956232H","6968816C","6935380E","6851184E","6950169G","6857152G","6855137S")
#this was achieved by viewing clabna.

#cplab <- cplab %>% filter(mrn %in% reassign_group_70p)
cplab[cplab$mrn %in% reassign_group_70p,]$group <- "apt"
```


```{r now lets use df cplab to calculate first in range values.}
dfm2 <- cplab %>% 
    filter((group == "axa" & axarange == "in")|(group == "apt" & apttrrange == "in")) %>% 
    group_by(mrn) %>% 
    slice_min(dtm)%>% 
    ungroup()

gh <- setdiff(pt$mrn,dfm2$mrn)
#why is this 34 patients not counted in dfm2 ?
#out of this group, most of them dont have axarange (is either below) and most of their apttrange is above or below.
#however this group have in range in aptr though they are in axa group.
z <- c("6853799C","6802209G","6858425U","6858634F")
#seems like these patients never got in range for AXA.

#now further lookign at it. dfm2 has a duplicated patient with a duplicated time stamp.
#thus lets remove the one where apttrrange is NA.
dfm2 <- dfm2[!((dfm2$mrn == "6941358K") & (is.na(dfm2$apttrrange))),]
```

```{r calc_ttinr}
#now let us calculate time to first in range.
#given how all of this would be first time in range

dfm2$fin <- dfm2$dtm - dfm2$datecannulated
dfm2$fin <- as.numeric(dfm2$fin)

#why is this much worse!
```



- Missing anti Xa results 
- Here again implicit and explicit.
- implicit missingness is where you dont measure it so it is not counted in your data. But in reality, it is missing.
-

103375H
2020-04-01 08:12:00

NB: we will use RLE and length to get drug prescription changes.

# Time to first anticoagulation analysis

let's make a survival object for time to first (tfirst), event outcome as (yes = achieving in range)

```{r}
sptf <- survfit (Surv(tfirst,event == "yes")~g2, data = dfg2p)

p13 <- ggsurvplot(
        sptf,
        pval = TRUE,
        fun = "event",
        risk.table = F, 
        risk.table.col = "g2",
        risk.table.height = 0.3,
        surv.median.line = "hv") + 
        labs (title = "Time to First Therapeutic Range", ylab = "% in Therapeutic Range",xlab = "Time/hours")

p13
```



# Time in therapeutic range.

#First Using traditional approach.
```{r testcode}
clab %>% filter (mrn %in% c("5743542Q","5764499G")) %>% group_by(mrn,axarange,apttrrange) %>% summarise(rt = min(ecmorunt), daysinr = n())
```

let's use g2 as is a more sophisticated definition. 
```{r}
cttraxa <- clab %>% 
          select(mrn,axarange,g2,cohort) %>% 
          filter(g2 == "axa")%>%  
          group_by(mrn) %>% 
          count(axarange) %>% 
          ungroup()

cttraxa <- left_join (cttraxa, clabna %>% select(mrn,erunt,no_lab),by = "mrn")

cttraxa <- cttraxa %>% 
            filter(axarange == "in") %>%
            group_by(mrn) %>% 
            mutate(tradttr = n/no_lab) %>%
            ungroup()
```

By using defintion 2 for grouping , the group breakdowns apt 81 and axa 120.
out of this 120 in axa, only 106 ever achieve in range. 
out of this 81 in apt, only 61 achieve in range. 

Comparing this, there is some signal 

let's do the same for apttr. 
```{r}

cttrapt <- clab %>% 
          select(mrn,apttrrange,g2,cohort) %>% 
          filter(g2 == "apt")%>%  
          group_by(mrn) %>% 
          count(apttrrange) %>% 
          ungroup()

cttrapt <- left_join (cttrapt, clabna %>% select(mrn,erunt,no_lab),by = "mrn")

cttrapt <- cttrapt %>% 
            filter(apttrrange == "in") %>%
            group_by(mrn) %>% 
            mutate(tradttr = n/no_lab) %>%
            ungroup()

```

#intsect(cttraxa$mrn,cttrapt$mrn)
showed 0. so no overlap. 

```{r}
cttraxa$g <- "axa"
cttrapt$g <- "apt"

cttr <- rbind (
        cttraxa %>% select(mrn,g,tradttr),
        cttrapt %>% select(mrn,g,tradttr)
)

cttr$g <- as.factor(cttr$g)
```

```{r}
p14 <- ggplot(data = cttr, aes(x = g,y = tradttr,colour = g)) +
      geom_boxplot()+
      theme_minimal()+
      labs (x = "Monitoring Group", 
            title = "Time in Therapeutic Range Traditional Method", 
            y = "No in range tests/Total no of tests", 
            subtitle = "106/120 achieved in range in AXA vs. 61/82 in APT. 
            Wilcox.test p-value = 0.006
            Assumptions = effects same over time")
     

p14
```


# Analysis questions

- does TTR varies over time? are we getting better or worse?
- does mortality risk varies over time? 
- does time on ecmo varies over time?

# Future Steps

- Is thrombotic complications routinely assesed and documented? - If so this will be useful for analysis.
- Is complications as per ELSO ? 
- Missing data in prescriptions and complications needs to be cross-checked 
- Obtain range to 31-Dec-2021 to include further data on covid?
- Further variables on patient(e.g., ethnicity, gender, kidney dysfunction, concurrent medications such as aspirin, clorpidorel, past medical history, SOFA score)
- Recommend a certain method in ECMO. i.e., the method that has least complications.
- Bland-Altman plot to stamp out bias in underlying measurement methods.
- Try at Brompton data?
- Change in circuits ? e.g., Heparin bonded circuits or MAHAMA/NO-doped circuits.
- frequency of ecmo circuit changes?
- platlet count, haemoglobin,CRP, LFT 




so units need deciding ? units vs. units/hr vs units/kg/hr
so kg needs deciding as "NA" vs "TRUE


may be that it is not NA but if you generate a time stamp for each time blood is monitored, then those with less frequent will have more NAs?


6956232H

6956232H
ambition

chep %>% group_by(mrn) %>% summarise(changes = length(rle(t_form)$lengths))%>% arrange(changes)%>%View()
