---
title: "Anti Xa Monitoring : A Preliminary Analysis"
author: "teddyhla"
date: "25/01/2022"
output: 
  html_document:
    toc : true
    toc_float : true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warnings = F)
#this is a global chunk options.
#for release - echo should be FALSE which means results will only appear. messages and warnings should also be false. 
```

<style>
@import url('https://fonts.googleapis.com/css2?family=Roboto&display=swap" rel="stylesheet');
body {
font-family: 'Roboto', sans-serif;
}
</style>


# INTRODUCTION

- Any extra-corporeal therapy requires some form of anticoagulation to reduce thrombotic complications. 
- Anticoagulation in critically unwell patients is challenging with rapid changes in physiology, biochemistry and disease progress.
- Several anticoagulants are available but intravenous unfractionated heparin is most commonly used.
- Modalities to monitor heparin exist from long-term monitoring (e.g., INR for warfarin monitoring) or from static coagulation tests.[cite]
- APTT test - activated partial thrombplastin time (aPTT) assay is a standard modality, available since the 1950's.
- APTT test is susceptible to interference from factor VIII or fibrinogen concentrations.
- Anti Xa assay - is thus a more attractive option as is not susceptible to factor deficiencies except anti-thrombin deficiency.
- Whilst anti Xa assay is recommended for monitoring by American College of Chest Physicians, College of American Pathologists, there is no definitive study showing the superiority of anti-Xa. 

# OBJECTIVES

- Does the use of anti-Xa monitoring results in 'better' outcomes?
- Definition of 'better' :
| Better efficacy - results in improved time in therapeutic range?
| Better patient outcomes - results in safer (i.e., less complications) in patients?
| Better provider outcomes - less monitoring frequency and drug dosage changes, potential for time-saving, cost, safety?

# AMBITION

- monitor all critically unwell patients with Anti-Xa ?

- development of an applet to guide dosage and monitoring?

- identification of variables associated with complications / worse outcomes? 

# PILOT STUDY : Dose the use of Anti-Xa monitoring in VV-ECMO patients improve outcomes?

- Explicit Inclusion criteria : all patients on VV-ECMO for viral respiratory failure from 01-01-2015 to 01-03-2021.

| Implicit Assumptions : all adults, predominant single organ failure of similar viral aetiology. 

- Timeframe : 1st November 2019 GSTT ECMO service switches to anti-Xa monitoring. 

- Methodology : Retrospective cohort study of all eligible patients, electronic health records interrogation to yield
1. patient's on ECMO run time 
2. Measured aPTTR and AXA levels whilst patients on ECMO
3. Anticoagulant prescriptions whilst on ECMO
4. Complications whilst on ECMO 

# SOURCE OF DATA

## Cohorts 

Three cohorts of severe respiratory failure patients are identified :- 
- wave 1 of covid - 19, defined as between 01-03-2020 to 31-08-2020 inclusive.
- wave 2 of covid - 19, defined as between 01-09-2020 to 01-03-2021 inclusive.
- historical flu, defined as 01-01-2015 and 31-12-2019 inclusive. 

## Questions regarding data.

- What about non covid and non flu patients in this period? isnt it better to capture 01-01-2015 to 01-01-2022 ? 
- Ans : better to focus on viral aetiology. 

- What about large gaps - is it true?
- Ans : as per above 

- What about lmwh / flolan / argotraban ? 
- Ans : only heparin was used 

- Use ecmo run time as gold standard? 
- Ans : agreed 

## Sensecheck 

- No repeat admissions?
- Ans : true

- Peak 3 ecmo admission in 24 hour ?
- Ans : true

## Patient identification
- ecmo blood pump speed was calculated from min to max dates and subtracted to get ecmo run times and patients

## Heparin administration

Heparin was administered in 4 different formats 

- heparin units/hr
- heparin ecmo xa iv infusion units/kg/hr
- heparin iv injection units
- heprin rrt inf 2000 iu units
- heparin systemic inf 2000 units units/hr

- Ans : agree this is the complexity of heparin prescription in ecmo.

### Heparin Protocol

Intravenous unfractionated heparin infusion (UFH) -  50 IU/kg as a bolus at the
time of cannulation with 2,500 units in the priming fluid for the circuit unless contraindicated
due to bleeding.


Prior to 1st November 2019, UFH infusions were monitored by activated partial thromboplastin
time ratio (APTTr) with a target level of 1.5-2.0 if evidence of no thrombosis and haemorrhage
and 2.0-2.5 if a thrombotic event.


From 1st November 2019 to present, anti-Xa activity was used for monitoring targeting levels
of 0.3 - 0.7 unit/mL if no evidence of thrombosis or haemorrhage were present and 0.6 - 1.0
unit/mL if there was a thrombotic ev

Weight is also extracted to calculate units/hr/hr 

## Coagulation lab results
- anti Xa and APTTR results were extracted

## Complications 

- Bleeding scores are extracted from EHR. ?Is this **ELSO** definitions?
- Thrombotic complications **NOT** available.


# DATA ANALYSIS

## Initialising 
### Load dependencies

```{r loading_libs}
library(ggplot2)
library(cowplot)
library(survminer)
library(survival)
library(plotly)
library(kableExtra)
```

### Source the cleaned files
```{r loading_source, message=TRUE,warning=FALSE}
source("datatidy.R")
```

## Description of Patient Population 

### Gross Statistics 
```{r baseline_table_01}
pt %>% 
        select(cohort,survivedecmo,ecmorunt) %>%
        group_by(cohort) %>% 
        summarise(
          no_pts = n(),
          no_deaths_ecmo = sum(survivedecmo == "no"),
          crude_mortality_rate = round(no_deaths_ecmo/no_pts,digits = 2),
          mean_ecmorunt = round(mean(ecmorunt),digits = 2),
          median_ecmorunt = median(ecmorunt),
          min_ecmorunt = min(ecmorunt)
                  ) %>%
        kbl() %>% 
        kable_styling()
        
```


### Admission Frequency of ECMO patients

```{r admission_frequency_cohort}
p2 <- ggplot(data = pt, aes(x = as.Date(admissiondate),colour = cohort)) +
      geom_bar()+
      scale_x_date(date_breaks = "6 months",date_minor_breaks = "3 months",date_labels = "%b %Y")+
      theme(axis.text.x = element_text(angle = 45,hjust = 1 ))+
      labs (x = "Time", title = "ECMO admissions across time")

p2
```


### Distribution of ECMO run times.

```{r hist_ecmo_runtimes}
p1 <- ggplot(data = pt, aes(x=as.numeric(ecmorunt), colour = cohort)) +
   geom_freqpoly(binwidth = 5 ) +
   scale_x_continuous(breaks = seq(0,90,by=10))+
   labs (x = "ECMO run time / days", y = "Patients", title = "Frequency of ECMO run times per cohorts")

ggplotly(p1)
```

### Dose age and cohort effect ECMO survival? 
```{r Age distribution across cohorts}

p4 <- ggpubr::ggboxplot(pt,x = "cohort",y = "age",fill ="cohort")
p4 + stat_compare_means()+ labs(title = "Age distribution across cohorts")
```

```{r cohort_outcomes}
p5 <- ggplot(data = pt, aes(x = cohort, fill = survivedecmo))+
    geom_bar(position = "stack") + 
    labs(x = "Cohorts", y = "Patients",title = "Patients who survived ECMO per cohort", subtitle = "No statistical difference")

p5 
```


### Time on ECMO to outcome 

```{r survival data}
ps <- survfit (Surv(ecmorunt,survivedecmo == "no")~cohort, data = pt)

p6 <- ggsurvplot(
        ps,
        pval = TRUE,
        risk.table = F, 
        risk.table.col = "cohort",
        risk.table.height = 0.3,
        surv.median.line = "hv") + labs (title = "Time on ECMO to outcome",x = "ECMO run time / days",y = "% on ECMO")

p6
```

### Risk Table for ECMO Run Times

```{r}
t1 <- ggsurvtable(ps,color = "cohort", data = pt,title = "% on ECMO over time in days")
t1$risk.table

```


so units need deciding ? units vs. units/hr vs units/kg/hr
so kg needs deciding as "NA" vs "TRUE


may be that it is not NA but if you generate a time stamp for each time blood is monitored, then those with less frequent will have more NAs?




ambition

chep %>% group_by(mrn) %>% summarise(changes = length(rle(t_form)$lengths))%>% arrange(changes)%>%View()
