---
title: "prelim_analysis"
author: "teddyhla"
date: "10/05/2022"
output: html_document
---

# Initialisation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r sources}
source("scripts/3-outcomes.R")
source("scripts/1c-clean.R")
```

```{r wrangle_core}
#duration on ecmo in hours
dfcore$ecmoh <- round(difftime(dfcore$ecmo_finish,dfcore$ecmo_start),1)

#duration on ecmo in days

dfcore$ecmod <- round(difftime(dfcore$ecmo_finish,dfcore$ecmo_start,units = "days"))

#assigning group 
dfcore$group <- ifelse(dfcore$date_can < as.Date("2019-12-01"),"gapt","gaxa")
dfcore$group <- as.factor(dfcore$group)
```

# Available Data

-   254 patients received veno-venous extracorporeal membrane oxygenation (VV-ECMO) at Guy's and St Thomas' Hospital NHS Trust (GSTT).
-   Admission to GSTT : 2nd January 2015 to 31st January 2022.

## Inclusion Criteria

-   all VV ECMO patients on viral aetiology

## Exclusion Criteria

# Variables available

## Cross-sectional

1.  Age - integer
2.  Admission date - date
3.  Cohort - factors; covid wave 1,2,3 and historical pneumonia
4.  diagnosis category - factor ; viral pneumonia
5.  mode of ecmo - factor ; vv
6.  ecmo cannula configuration - factors; access and return
7.  survival on ecmo - binary factor ; yes/no
8.  survival from icu - binary factor ; yes/no
9.  date of icu discharge
10. ecmo start and finish dates
11. date of ecmo cannulation and date of ecmo de-cannulation
12. acute physiology score
13. age score
14. chronic health score
15. apache score
16. ethnicity
17. gender
18. height
19. weight
20. past medical history - factors; cirrhosis, congential immunodeficiency, hiv, steroids history and metastatic

## Longitudinal

Data available for time and values

21. complete blood count parameters (haemoglobin,platelets,neutrophils)

22. coagulation parameters (fibrinogen,)

23. inflammatory parameters (ldh,ferritin,creatinine kinase,crp, procaclitonin)

24. synthetic liver function (bilirubin,albumin,creatinine)

25. renal function tests (gfr,creatinine)

26. electrolytes(calcium, corrected_calcium)

27. acid_base parameters( bicarbonate, ph,lactate )

28. anti factor Xa levels (axa)

29. activated partial thromboplasmin time ratio (apptr)

30. heparin prescription - factors

31. heparin dosage - weight adjusted

32. hydrocortisone infusions

33. blood product transfusions (cryoprecipitate, ffp, platelet, packed red cells)

34. other drugs (aspirin,aspirin suppository,clopidogrel,warfarin, rivaroxaban,apixaan, dalteparin subcutnaeous, ibuprofen liquid, ibuprofen tablet, diclofenac, dexamethasone, prednisolone, hydrocortisone, methylprednisolone, calciu, tranexemic acid,tranexemic acid infusions )

35. admission CT findings

36. interval CT findings

37. additional complications haemorrhagic / thrombotic

38. decannulation ultrasound findings

39. ecmo circuit change and reasons

# Analysis

-   joint models
-   cox Proportional Hazards model
-   Poisson regression

# Challenges

## Different Targets

### PE : arterial thrombus?

patients with PE will have a separate target other anticoagulants - e.g., ischemic leg need to check 6 patients on PE

questiosn for silvana 2639533Z - covid pneumonitis with ischemic leg - is treated as PE. i.e., arterial thrombus

### Heparin Prescriptions

[1] "Heparin" "Heparin ECMO Xa IV INFUSION"\
[3] "Heparin IV Injection" "Heparin RRT INF 20000 iu"\
[5] "Heparin SYSTEMIC INF 20000Units"

making sense of these prescriptions

                        Heparin     Heparin ECMO Xa IV INFUSION            Heparin IV Injection 
                             41                           57889                              39 
       Heparin RRT INF 20000 iu Heparin SYSTEMIC INF 20000Units 
                             17                           33617 

### heparin free

17 patients

# Findings

## Ecmo configuration

-   there is no 2nd access.
-   there is no arterial return

## Group assignment

1st November 2019 is when AXA monitoring is actively initiated. But people dont formally start using until 2019-12-01

## Descriptive Statistics

### Group Comparison

```{r group_comparison_table}
dfcore %>% 
    group_by(group) %>% 
    summarise(
        medage = median(age),
        iqrage = IQR(age),
        counts = n(),
        surv_ecmoy = sum(surv_ecmo == "yes"),
        surv_ecmon = sum(surv_ecmo == "no"),
        runtd = median(ecmod),
        iqrtd = IQR(ecmod),
        medapache = median(apache),
        iqrapach = IQR(apache),
        medweight = median(wkg),
        iqrweight = IQR(wkg),
        sexm = sum(sex == "M")
        
    )

```

### Stats for above group comparison

```{r stats_comp_group}
xs1 <- chisq.test(dfcore$age,dfcore$group)
xs2 <- chisq.test(dfcore$sex,dfcore$group)
wc1 <- wilcox.test(apache ~ group, data =dfcore)
wc2 <- wilcox.test(wkg ~ group, data =dfcore)
xs3 <- chisq.test(dfcore$surv_ecmo,dfcore$group)
```

```{r plotforgroupage}
ggplot(data =dfcore, aes(x=group,y=ecmod)) + geom_boxplot()

```

that kind of made sense

```{r}
dfbl <- left_join (
    dfbl,
    dfcore %>% select(mrn,ecmo_start,ecmo_finish,group),
    by = "mrn"
)

dfbl$chart_t <- as.Date(dfbl$chart_t,format ="%Y-%m-%d") 
dfbl$ecmo_start <- as.Date(dfbl$ecmo_start,format = "%Y-%m-%d")
dfbl$ecmodur <- dfbl$ecmod
dfbl$decmo <- difftime(dfbl$chart_t,dfbl$ecmo_start,units = "days")
dfbl$decmo <- as.numeric(dfbl$decmo)
dfbl$decmo <- dfbl$decmo + 1

dfbl$thours <- difftime(dfbl$chart_t,dfbl$ecmo_start,units = "hours")
dfbl$tdays <- round(difftime(dfbl$chart_t,dfbl$ecmo_start,units = "days"))


dfhb <- dfbl %>% select(mrn,group,hb,thours,tdays)
dfhb <- na.omit(dfhb)
```

'

```{r}


dfhb2 <- dfhb %>% 
    group_by(mrn,tdays) %>% 
    mutate(
        maxhb = max(hb,na.rm = TRUE),
        minhb =min(hb,na.rm=TRUE),
        meanhb=mean(hb,na.rm=TRUE)) %>%
    ungroup()
```

```{r}
#dfhb2 %>% group_by(group,mrn) %>% mutate(perc = hbhi/ecmod) %>% ungroup () %>% #distinct(mrn,.keep_all = TRUE)
#you want something like this.
```

```{r}
#dtf %>% 
#    group_by(group) %>% 
#    group_map(~summary(.x))
    
```

```{r}
dfprd <- left_join(
    dfprd,
    dfcore %>% select(mrn,ecmo_start,ecmo_finish,group,ecmod),
    by = "mrn"
)

tfpp <- dfprd %>% 
    group_by(mrn) %>% 
    summarise(
        prbcpp = sum(s_label == "prbcs"),
        cryopp = sum(s_label == "cryo"),
        ffppp = sum(s_label == "ffp"),
        pltpp = sum(s_label == "plt")
    )

tfpp <- left_join(
    tfpp,
    dfcore %>% select(mrn,ecmod,group),
    by = "mrn"
)

tfpp$ecmod <- as.numeric(tfpp$ecmod)
```

```{r}
tfpp$redpd <- (tfpp$prbcpp / tfpp$ecmod)*1000
tfpp$crpd <- (tfpp$cryopp / tfpp$ecmod)*1000
tfpp$fpd <- (tfpp$ffppp / tfpp$ecmod)*1000
tfpp$ppd <- (tfpp$pltpp / tfpp$ecmod)*1000

#becuase it gets too small

tfpp %>% group_by(group) %>% group_map(~summary(.x))

```

```{r}
dfcore %>% filter(mrn %in% pt_without_bldproducts) %>% group_by(group) %>% summarise(n())
```

let's look at descriptive analysis of outcomes

```{r}
df <- left_join(
    df,
    dfcore %>% select(mrn,group),
    by = "mrn"
)

```

```{r}
po1 <- ggplot(data = df,aes(x=tot,fill = group)) +
           geom_bar(position = "dodge")+
           labs(title = "Total No of Complications ", 
             x = "Total No of Complications", y = "Count")+
           theme_minimal()
```

```{r}
ggsave("products/presentations/gsttp2src/go1.png", po1, dpi = 300)
```

# Things to consider

1.  cant use axa values in the apttr group
2.  drugs

-   no warfarin, no rivaroxabab,no apixaban, diclofena inj or suppr, hydrocort tab, txa top

# for presentation

## Playing with alluvial plots(rb

```{r p1 age_distribution}
p1 <- ggplot(data = dfcore, aes(x = cohort, y = age))+
        geom_boxplot()+
        facet_wrap(~surv_ecmo)+
        theme_minimal() + 
        labs(title = "Age distribution for each cohort", 
             x = "cohorts", y = "age/years",
             subtitle = "surival status on ecmo")+
        theme(axis.text.x = element_text(angle = 45))
        
```

```{r savingp1}
ggsave("products/presentations/gsttp2src/g1.png", p1, dpi = 300)

```

```{r attempt at alluvial plots }
#al <- df %>% select(admn_ct,int_ct,decan_uss) %>%
#        group_by(admn_ct,int_ct,decan_uss) %>%
#        summarise(freq = n())
#
#p2 <- ggplot(data = al, aes(axis1 = admn_ct, axis2= int_ct, axis3 = decan_uss,
#                            y = freq))+
#        scale_x_discrete(limits = c("admn_ct","int_ct","decan_uss"))+
#        xlab("imagings") +
#        geom_alluvium()+
#        geom_stratum()+
#        geom_text(stat = "stratum",aes(label=after_stat(stratum)))+
#        ggtitle("Patients who received imagings whilist on ecmo")
#        
```

For outcomes, alluvial cant handle missing data.

```{r}
#al2 <- df %>% select(admn_ct_results,int_ct_results,total_events_not_imaged,decan_uss_results#)
#
#al2$admn_ct_results <- as.character(al2$admn_ct_results)
#al2$int_ct_results <- as.character(al2$int_ct_results)
#al2$total_events_not_imaged <- as.character(al2$total_events_not_imaged)
#al2$decan_uss_results <- as.character(al2$decan_uss_results)
#
#al2[is.na(al2)] <- 0
#
#al2$admn_ct_results <- as.factor(al2$admn_ct_results)
#al2$int_ct_results <- as.factor(al2$int_ct_results)
#al2$total_events_not_imaged <- as.factor(al2$total_events_not_imaged)
#al2$decan_uss_results <- as.factor(al2$decan_uss_results)
#
#levels(al2$int_ct_results)<- c(
#        "0",
#        "both_haem_throm",
#        "only_haem",
#        "no_comp",
#        "no_comp",
#        "only_haem",
#        "only_throm",
#        "only_throm"
#        )
```

```{r}
#al2 <- al2 %>% group_by(admn_ct_results,int_ct_results,total_events_not_imaged,decan_uss_res#ults)%>%
#        summarise(freq = n())
```

```{r}
#p3 <- ggplot(data = al2, #aes(axis1=admn_ct_results,axis2=int_ct_results,axis3=total_events_not_imaged,axis4=decan_uss_results,y=freq))+
#        scale_x_discrete(limits = #c("admn_ct_results","int_ct_results","other_events","decan_uss_results"))+
#        geom_alluvium()+
#        geom_stratum()+
#        geom_text(stat = "stratum",aes(label = after_stat(stratum)))+
#        ggtitle("complications whilst on ECMO")
```

```{r}
#ggsave("products/presentations/gsttp2src/g3.png",p3,dpi = 300)
```

## Raw distributions and Standardised raw visualisations

```{r}
vin <- dfcoag
vin <- left_join(vin, 
                 dfcore %>% select(cohort,group,ecmo_start,ecmo_finish,mrn),
                 by ="mrn")

vin <- vin %>% 
    group_by(mrn) %>% 
    filter(chart_t >= ecmo_start & chart_t <= ecmo_finish) %>% ungroup()

vin$ecmod <- round(difftime(vin$chart_t,vin$ecmo_start, units = "days"))
vin <- vin %>% filter(ecmod >= 0)

#0.5 for mid point of axa , sd axa = 0.280/mean is 0.51 and 1.75 for apttr, calc mean is 1.58, sd is 1.604

#thus manually scaling i.e., minu by midpoint and divide by SD
vin$axas <- ((vin$axa - 0.5)/0.280)
vin$apttrs <- ((vin$apttr - 1.75/1.604))

vin$ecmod <- as.factor(vin$ecmod)

#ideally select during the ecmo run's


```

```{r}
ylim1 <- boxplot.stats(vin$axa)$stats[c(1,5)]
ylim2 <- boxplot.stats(vin$apttr)$stats[c(1,5)]


p7<-ggplot(data = vin[vin$group == "gaxa",], aes(x = ecmod, y = axa)) +
    geom_boxplot(outlier.shape = NA)+
    geom_hline(yintercept = c(0.3,0.7),colour = "blue")+
    geom_hline(yintercept = 1.0,colour ="red")+
    coord_cartesian(ylim = ylim1 *1.05)+
    scale_x_discrete(breaks = seq(0,80,10))+
    theme(axis.text.x = element_text(angle = 45))+
    theme_minimal()+
    labs(title= "AXA values in AXA group",
         x = "Days on Ecmo",y ="anti-Xa levels",
         subtitle = "Normal Range 0.3 to 0.7, VTE Range 0.6 to 1.0")


p8<- ggplot(data = vin[vin$group == "gapt",], aes(x = ecmod, y = apttr)) +
        geom_boxplot(outlier.shape = NA)+
        geom_hline(yintercept = c(1.5,2.5),colour = "blue")+
        geom_hline(yintercept = 3.0,colour = "red")+
        coord_cartesian(ylim = ylim2 *1.05)+
        scale_x_discrete(breaks = seq(0,80,10))+
        theme(axis.text.x = element_text(angle = 45))+
        theme_minimal()+
        labs(title= "APTTR values in APTT group",
         x = "Days on Ecmo",y ="APTTR values",
          subtitle = "Normal Range 1.5 to 2.0, VTE Range 2.0 to 2.5")
```

```{r}
p9 <- cowplot::plot_grid(p7,p8)
```

```{r}
ylim1 <- boxplot.stats(vin$axas)$stats[c(1,5)]
ylim2 <- boxplot.stats(vin$apttrs)$stats[c(1,5)]

p10<-ggplot(data = vin[vin$group =="gaxa",], aes(x = ecmod, y = axas)) +
    geom_boxplot(outlier.shape = NA)+
    geom_hline(yintercept = c(-1,1),colour = "green")+
        coord_cartesian(ylim = c(-2,2))+
    scale_x_discrete(breaks = seq(0,80,10))+
    theme(axis.text.x = element_text(angle = 45))+
    theme_minimal()+
    labs(title= "SCALED AXA values in AXA group",
         x = "Days on Ecmo",y ="scaled AXA values")


p11<- ggplot(data = vin[vin$group == "gapt",], aes(x = ecmod, y = apttrs)) +
        geom_boxplot(outlier.shape = NA)+
    geom_hline(yintercept = c(-1,1),colour = "green")+
        coord_cartesian(ylim = c(-2,2))+
    scale_x_discrete(breaks = seq(0,80,10))+
        theme(axis.text.x = element_text(angle = 45))+
        theme_minimal()+
        labs(title= "SCALED APTTR values in APTTR group",
         x = "Days on Ecmo",y ="scaled APTTR values")
```

```{r}
p12 <- cowplot::plot_grid(p10,p11)
```

```{r}
ggsave("products/presentations/gsttp2src/g4.png",p9,dpi = 300)
ggsave("products/presentations/gsttp2src/g5.png",p12,dpi = 300)
```

## are the patients proportionaley distributed into cohorts

```{r}

```

## how many non covid patients are in each cohort? there are none

```{r}
#df %>% select(cohort,dxcont) %>% group_by(cohort,dxcont) %>% summarise(counts = n())
```

##let's explore time variation of ecmo outcomes and durations

lets append to

```{r}
df <- left_join(
    df,
    dfcore %>% select(mrn,ecmo_start,ecmo_finish,ecmod,ecmoh),
    by = "mrn"
)
```

Basically the collected df ecmo_dtm are not accurate. we will remove them

```{r}
df <- df %>% select(-c(time_ecmo_DE_can,ecmo_dtm,ecmo_end_dtm))

```

```{r}
df$admnct_
```

we should make a big data table age,gender,bmi,past medical history,dx cat, hb, plt,neut, fib, ldh. ferritin, ck, crp, pct, bili, alb, creat lactate,ph

#Lets look at heparin prescriptions

```{r}
dfhep <- left_join (
    dfhep,
    dfcore %>% select(mrn,group,ecmod), by = "mrn"
)
```

# Poisson regression for transfusion

vs. Negative binomial model

## explore only RBC vs. all products.

## what about fibrinogen?

think about begininig nd end of ecmo run. think about other complications e.g., haemorrhagic complications

#deciding which feature to add - day on ecmo . hypothesis = we do need to as we transfuse more at the begining.

#lets look at total products for ecmo run

note median ecmo run time = 14 days 1st quartile 8 days 2nd quartile 14 days 3rd quartile 25.75 days 4th quartile 100 days

```{r}
e2 <- dfprd %>% 
        group_by(ecmod) %>%
        summarise(
                totcryo = sum(s_label == "cryo"),
                totffp = sum(s_label == "ffp"),
                totplt = sum(s_label == "plt"),
                totprbc = sum(s_label == "prbcs")
        )

e2$ecmod <- as.numeric(e2$ecmod)
e2 <- e2 %>% mutate(ecmodq = 
        case_when(
                ecmod <= 8 ~ "q1",
                ecmod > 8 & ecmod < 14 ~ "q2",
                ecmod > 14 & ecmod <25.75 ~ "q3",
                ecmod > 25.75 & ecmod < 100 ~ "q4"
        )
)
```

Although this doesnt mean anything for each individual patient.

for each individual patient, what you want is 4 segments !

What is an average time between each red cell transfusions

```{r}


e4 <- dfprd %>% 
        filter(s_label == "prbcs") %>% 
        group_by(mrn) %>% 
        arrange(chart_t) %>% 
        group_by(mrn) %>% 
        mutate(tdi = difftime(chart_t,lag(chart_t),units = "days"))%>% 
        ungroup()

e4 %>% group_by(group) %>% summarise(quantile(tdi,na.rm = T))
#showing most patients receive a transfusion in 24 hour sort of cycle. 
```

thus we can set 1 day as a unit of 'window' tranfusion in this window "yes" "no"

may be we should look at major hemorrhage events

```{r}
dfprd %>% group_by(mrn,td) %>% summarise(rpd = sum(s_label == "prbcs")) %>% ungroup() %>% group_by(mrn) %>% arrange(desc(rpd),.by_group = TRUE) %>% View()
```

what count as major hemorrhage. what coutns as massive transfusion

seems like we have a break through.

let's make a dataframe that has all the days on ecmo including day 0

so this below code works. now lets append this with total no of products transfused

```{r}
dg <- dfcore %>% select(mrn,group,ecmod) %>% mutate(ecmod = as.numeric(ecmod)) %>% group_by(mrn) %>% uncount(weights = ecmod, .id = "ecmod")

#lets make a dataframe of summed blood products

dg1 <- dfprd %>% 
        select(mrn,group,td,s_label,t_form) %>%
        mutate(td = as.numeric(td)) %>%
        group_by(mrn,td) %>% 
        summarise(
                totprbc = sum(s_label == "prbcs"),
                totffp  = sum(s_label == "ffp"),
                totcryo = sum(s_label == "cryo"),
                totplt = sum(s_label == "plt"),
                totall = totprbc + totffp + totcryo + totplt
        ) %>%
        ungroup() 

#for th time being lets remove random outiers where 
dg1 <- dg1[dg1$td > 0 & dg1$td <88,]

```

needs to clarify with silvana

lets merge it all together

```{r}
dg$mid <- paste(dg$mrn,dg$ecmod,sep = "")
dg1$mid <- paste(dg1$mrn,dg1$td,sep = "")

dg <- left_join (
        dg,
        dg1,
        by = "mid"
)

dg <- dg %>% 
        mutate(across(where(is.numeric),~replace_na(.x,0))) %>%
        select(-c(mrn.y,td,mid))

names(dg)[names(dg) =="mrn.x"]<- "mrn"
```

after all this total blood units transfused should check out but it is not

following is a code for looking at patients who have NA in the t_form

#We decided not to have 0th day on ecmo. we will have 1st day and nth day on ecmo.

```{r}
dfprd[is.na(dfprd$t_form),] %>% group_by(mrn)
```

how do we analyse day 1? do we analyse from day 2? may be we cant analyse that.

```{r}
dfbl <- dfbl %>% 
        select(- c(ecmo_start,ecmo_finish,thours)) %>% 
        mutate(tdays = as.numeric(tdays)) %>% 
        mutate(ecmod = tdays + 1)
```

current methdo of setting dates not so robust for dfbl especially. we need to revise.

on average how many days do the patient got transfused. seems not every day

```{r}
dtxprd <- dg %>% 
        group_by(mrn) %>%
        summarise(
                maxd = max(ecmod),
                dtx = sum(totall != 0 )
        ) %>%
        mutate(prpn = dtx/maxd)
```

This code looks at number of days where patients are transfused (dtx) out of their total days(ecmod). The proportion (prpn) of 1 suggest daily transfusion. It suggests that only 6 patients receive daily transfusion All others have less than 0.5 Therefore it makes sense to use a 'day' as a unit.

Therefore we can use 1 day as a main unit

```{r}

```
